---
title: "Understanding Matrices and Logistic Regression in Neural Networks"
format: html
---
```{r}
#| echo: false
library(dplyr)
library(ggplot2)
library(tidyr)
theme_set(theme_minimal())
options(scipen= 999)
```

In this document, we explore the foundations of matrices in neural networks, logistic regression, forward propagation, and updating weights using stochastic gradient descent. We will implement a basic example in `R` to reinforce learning.

# Understanding Matrices in Neural Networks

Matrices enable efficient mathematical operations in neural networks. For a simple model:

$$
Z = XW
$$

where:
- $X$ is the **input matrix** (containing feature values).
- $W$ is the **weight matrix** (containing learned coefficients).
- $Z$ is the **output before activation**.

### **Traditional Logistic Regression Notation**
In standard **logistic regression**, the equation is expressed using individual predictor variables:

$$
Z = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_n x_n
$$

where:
- $\beta_0$ is the **intercept (bias term)**.
- $\beta_1, \beta_2, \dots, \beta_n$ are the **weights (coefficients) assigned to each feature**.
- $x_1, x_2, \dots, x_n$ are the **input feature values** (e.g., time spent, pages visited).
- $Z$ is the **linear combination before applying the activation function**.

Once $Z$ is computed, the logistic function (**sigmoid activation**) is applied:

$$
\sigma(Z) = \frac{1}{1 + e^{-Z}}
$$

Both forms describe the same underlying concept:
- The **matrix form** ($Z = XW$) is compact and efficient, especially when handling multiple samples.
- The **traditional logistic regression form** explicitly shows the relationship between individual features and their respective weights.

Both representations lead to the same outcome: a **probability prediction** via the sigmoid function.

# Logistic Regression Model
Logistic regression predicts probabilities using the **sigmoid function**:
The sigmoid function is a mathematical function that outputs values between 0 and 1, making it ideal for logistic regression, where we interpret the result as a probability.
$$\sigma(Z) = \frac{1}{1 + e^{-Z}}$$
where:
- $z$ is the input value (can be any real number)
- $e$ is Euler's number 

If $z$ is large and positive, $\sigma(z)$ approaches 1 (strong positive probability)
If $z$ is large and negative, $\sigma(z)$ approaches 0 (strong negative probability)
when $z=0$, $\sigma(z)=0.5)$, meaning neutral probability
```{r, echo=FALSE}
# Define the sigmoid function
sigmoid <- function(z) {
  return(1 / (1 + exp(-z)))
}

# Generate values for plotting
z_values <- seq(-10, 10, by=0.1)  # Range from -10 to 10
sigmoid_values <- sigmoid(z_values)

# Create a dataframe for plotting
df <- data.frame(z = z_values, sigmoid = sigmoid_values)

# Plot the sigmoid function
ggplot(df, aes(x = z, y = sigmoid)) +
  geom_line(color = "blue", size = 1) +
  ggtitle("Sigmoid Function") +
  xlab("z") +
  ylab("Ïƒ(z)") +
  theme_minimal()

```


# Forward Propagation & Loss Function
Predictions ($\hat{y}$) are made using:
$$\hat{y} = \sigma(Z)$$
The **binary cross-entropy loss function** quantifies the error between the estimate and the real output provided to the model:
$$L = -\frac{1}{m} \sum_{i=1}^{m} \left[ y_i \log(\hat{y}_i) + (1 - y_i) \log(1 - \hat{y}_i) \right]$$

# Updating Weights with Gradient Descent
Since we don't know the optimal values of $W$, we start with random weights and iteratively update them to minimize the loss function. Now, let's break down the equation further and clarify the partial derivatives.

Gradient loss function:
$$
\frac{\partial L}{\partial W}
$$
This term represents the gradient, or the slope, of the loss function $L$ with respect to $W$. Essentially, it tells us:

How much the loss function changes when we slightly change $W$
The direction we should move $W$ to minimize the loss.

Since $L$ depends on $W$ (because changing ð‘Šaffects predictions), we need to calculate the rate of change of $L$ concerning $W$, which is where the partial derivative comes in.

*Partial Derivatives Explained*
A partial derivative calculates how one variable changes while keeping others constant. In our case:
$$
\frac{\partial L}{\partial W}
$$
measures how much the loss function changes if we make a small adjustment to $W$.
If we visualize the loss function as a mountain, the gradient tells us which direction leads us downhill the fastest (toward lower loss).
The gradient gives the best direction for adjusting W, but how far we step in that direction is controlled by the learning rate $\alpha$
$$
W = W -\alpha \cdot \frac{\partial L}{\partial W}
$$
where:
- $ \alpha $ is the **learning rate** (a small step size to prevent large jumps)
- $ \frac{\partial L}{\partial W} $ is the **gradient**
$W$ is updated gradually with each iteration.

Fro logistic regression, the gradient of the *binary cross-entropy loss function* with respect to $W$ is:
$$
 \frac{\partial L}{\partial W} = \frac{1}{m}X^T(\hat y -y)
$$
Where:
- $X$ is the input matrix (features)
- $y$ is the actual target values.
- $\hat y = \sigma(XW)$ is the predicted output after applying the sigmoid function.
- $X^T(\hat y- y)$ measures the error's contribution to weight updates.

```{r}
# Example loss reduction over iterations
iterations <- seq(1, 100)
loss_values <- exp(-0.05 * iterations)  # Simulated loss decreasing

# Create dataframe
df <- data.frame(iteration = iterations, loss = loss_values)

# Plot loss reduction
ggplot(df, aes(x = iteration, y = loss)) +
  geom_line(color = "red", size = 1) +
  ggtitle("Loss Reduction Over Iterations") +
  xlab("Iteration") +
  ylab("Loss") +
  theme_minimal()

```


# Implementation in R

Scenario: Predicting Whether Someone Will Buy a Product
Imagine you're running an online store, and you want to predict whether a customer will buy a product based on two simple features:

- Time spent on the website (in minutes)
- Number of pages visited

We'll create a small dataset with these features and train a logistic regression model using gradient descent to predict whether a customer will buy the product (1) or not (0).

The traditional logistical regression formula would be:
$$
Z=\beta_0+\beta_1 x_{1i}+\beta_2 x_{2i}+\epsilon_i
$$

where:
- $\beta_0$ is the **intercept (bias term)**.
- $\beta_1, \beta_2$ are the **weights (coefficients) assigned to each feature**.
- $x_{1i}, x_{2i}$ are the **feature values** for the $i$-th sample.
- $\epsilon_i$ is the **error term** accounting for noise in the data.


Instead of writing the equation explicitly for each feature, we can use matrix multiplication:

$$
Z = XW + \epsilon
$$

where:
- $X$ is the **input matrix** containing feature values.
- $W$ is the **weight matrix** (vector of coefficients).
- $\epsilon$ is the **error term**.

Expanding this in **matrix notation**:

$$
\begin{bmatrix}
Z_1 \\
Z_2 \\
Z_3 \\
\vdots \\
Z_m
\end{bmatrix}
=
\begin{bmatrix}
1 & x_{11} & x_{21} \\
1 & x_{12} & x_{22} \\
1 & x_{13} & x_{23} \\
\vdots & \vdots & \vdots \\
1 & x_{1m} & x_{2m}
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\
\beta_1 \\
\beta_2
\end{bmatrix}
+ 
\begin{bmatrix}
\epsilon_1 \\
\epsilon_2 \\
\epsilon_3 \\
\vdots \\
\epsilon_m
\end{bmatrix}
$$

where:
- Each row in $X$ represents **one sample**, including a **bias term (1), feature 1, and feature 2**.
- The column vector $W$ contains the **learned parameters** ($\beta_0, \beta_1, \beta_2$).
- The error term $\epsilon$ accounts for **random noise in predictions**.


We have data from 5 customers, we will create a matrix for this:
Each row in $X$ represents a customer.
First column = Bias term (always 1).
Second column = Time spent on the website.
Third column = Number of pages visited.
y contains whether the customer bought the product (1) or not (0).

```{r}
 # Real-world dataset (Time spent & Pages visited)
X <- matrix(c(
  1, 5, 2,
  1, 15, 5,
  1, 20, 7,
  1, 2, 1,
  1, 30, 10
), ncol=3, byrow=TRUE)

y <- c(0, 1, 1, 0, 1)  # Labels
```

As a recapitulation, we use the sigmoid function to predict the expected output of any specific combination of minutes in the website and pages visited. We will create a function for this:

```{r}

# Sigmoid function
sigmoid <- function(z) {
  return(1 / (1 + exp(-z)))
}
```

the loss function will tell us how far our predicted value is from the actual value, we will also create a function for this:

```{r}
# Loss function (binary cross-entropy)
loss_function <- function(y, y_hat) {
  return(-mean(y * log(y_hat) + (1 - y) * log(1 - y_hat)))
}
```

Now we create a function to calculate the gradient descent and store the results of each iteration in a dataset so we can visualize it. 

```{r}
# Gradient Descent Implementation (Track Progress)
gradient_descent_progress <- function(X, y, learning_rate = 0.01, iterations = 50) {
  m <- nrow(X)
  W <- runif(ncol(X))  # Initialize weights randomly
  progress <- data.frame(Iteration = integer(), 
                         Weight1 = numeric(), Weight2 = numeric(),
                         Z1 = numeric(), Z2 = numeric(), Z3 = numeric(), Z4 = numeric(), Z5 = numeric(),
                         y_hat1 = numeric(), y_hat2 = numeric(), y_hat3 = numeric(), y_hat4 = numeric(), y_hat5 = numeric(),
                         Grad1 = numeric(), Grad2 = numeric())  # Tracking all five samples

  for (i in 1:iterations) {
    Z <- X %*% W   # Compute Z values for all samples
    y_hat <- sigmoid(Z)  # Apply sigmoid function to all samples
    gradient <- t(X) %*% (y_hat - y) / m  # Compute gradient for weight updates
    W <- W - learning_rate * gradient  # Update weights

    # Store results for all five samples
    progress <- rbind(progress, data.frame(
      Iteration = i,
      Weight1 = W[2], Weight2 = W[3],
      Z1 = Z[1], Z2 = Z[2], Z3 = Z[3], Z4 = Z[4], Z5 = Z[5],
      y_hat1 = y_hat[1], y_hat2 = y_hat[2], y_hat3 = y_hat[3], y_hat4 = y_hat[4], y_hat5 = y_hat[5],
      Grad1 = gradient[2], Grad2 = gradient[3]
    ))
  }
  
  return(progress)
}
```
At the beginning, the weights are random.

Over each iteration, the values change towards better predictions.
- Z values show the raw linear transformation before activation.

- y_hat tracks how probabilities evolve as weights adjust.

- Gradient values indicate how weights update to minimize the loss.

If you increase iterations, you'll see further refinement!
```{r}

# Train Model


# Run gradient descent and track evolution
progress_df <- gradient_descent_progress(X, y)

# Print evolution table
print(head(progress_df))

# Plot how weights change over iterations
ggplot(progress_df, aes(x = Iteration)) +
  geom_point(aes(y = Weight1, color = "Weight1"), size = 1) +
  geom_point(aes(y = Weight2, color = "Weight2"), size = 1) +
  ggtitle("Gradient Descent: Evolution of Weights") +
  xlab("Iteration") +
  ylab("Weight Value") +
  theme_minimal() +
  scale_color_manual(values = c("Weight1" = "blue", "Weight2" = "red"))


# Reshape progress_df for separate plots
long_df <- tidyr::gather(progress_df, key = "Variable", value = "Value", -Iteration)

# **Plot 1: Gradient Evolution**
gradient_df <- long_df %>% filter(Variable %in% c("Grad1", "Grad2"))

ggplot(gradient_df, aes(x = Iteration, y = Value, color = Variable)) +
  geom_point(size = 2) +  # Use points instead of lines
  ggtitle("Gradient Descent: Evolution of Gradients") +
  xlab("Iteration") +
  ylab("Gradient Value") +
  theme_minimal()

# **Plot 2: Predicted Probabilities (y_hat) Evolution**
y_hat_df <- long_df %>% filter(grepl("y_hat", Variable))

ggplot(y_hat_df, aes(x = Iteration, y = Value, color = Variable)) +
  geom_point(size = 2) +  # Use points instead of lines
  ggtitle("Gradient Descent: Evolution of Predicted Probabilities") +
  xlab("Iteration") +
  ylab("Probability (y_hat)") +
  theme_minimal()

# **Plot 3: Z Values Evolution**
z_values_df <- long_df %>% filter(grepl("Z", Variable))

ggplot(z_values_df, aes(x = Iteration, y = Value, color = Variable)) +
  geom_point(size = 2) +  # Use points instead of lines
  ggtitle("Gradient Descent: Evolution of Z Values") +
  xlab("Iteration") +
  ylab("Z Value") +
  theme_minimal()
```

**Understanding the Learned Weights**
After running gradient descent, we obtained the following learned weights:

the final weights:
```{r}
set.seed(2)
# Gradient Descent Implementation (Track Progress)
gradient_descent <- function(X, y, learning_rate = 0.01, iterations = 50) {
  m <- nrow(X)
  W <- runif(ncol(X))  # Initialize weights randomly
  for (i in 1:iterations) {
    Z <- X %*% W   # Compute Z values for all samples
    y_hat <- sigmoid(Z)  # Apply sigmoid function to all samples
    gradient <- t(X) %*% (y_hat - y) / m  # Compute gradient for weight updates
    W <- W - learning_rate * gradient  # Update weights
  }
  return (W)
}
W <- gradient_descent(X, y)
print(W)
```

These correspond to:
- $W_0$ â†’ **Bias term** (Intercept).
- $W_1$ â†’ **Effect of Time Spent on probability of purchasing**.
- $W_2$ â†’ **Effect of Pages Visited on probability of purchasing**.A positive value increases probability, while a negative value decreases probability.

To predict whether a new customer will buy a product, we use the following equation:

$$
Z = W_0 + W_1 \cdot \text{Time Spent} + W_2 \cdot \text{Pages Visited}
$$

Once we calculate $Z$, we apply the **sigmoid activation function**:

$$
\sigma(Z) = \frac{1}{1 + e^{-Z}}
$$

where $\sigma(Z)$ represents the **probability** that the customer will buy.


**Example Calculation**
Let's take a new customer who spends **12 minutes** on the website and visits **4 pages**. We calculate $Z$ as:

$$
Z = 0.01567795 + (0.09666904 \times 12) + (0.31485151 \times 4)
$$

Applying the learned weights, the prediction follows:

$$
Z = 2.435112
$$

Applying the sigmoid function:

$$
\sigma(2.435112) = \frac{1}{1 + e^{-2.435112}}
$$

Approximating:

$$
\sigma(2.435112) \approx 0.91
$$

Thus, the model predicts **91% probability** that this customer **will buy** the product.

let's see the predictions calculated in `r` over the same data we used for training:

```{r}
# Get predictions

# Function to predict probability based on learned weights
predict_probability <- function(X, W) {
  return(sigmoid(X %*% W))
}
predicted_probs <- predict_probability(X, W)

# Create dataframe for plotting
df <- data.frame(TimeSpent = X[,2], PagesVisited = X[,3], Probability = predicted_probs)

# Plot results
ggplot(df, aes(x = TimeSpent, y = PagesVisited, color = Probability)) +
  geom_point(size = 4) +
  scale_color_gradient(low = "blue", high = "red") +
  ggtitle("Logistic Regression: Probability of Purchase") +
  xlab("Time Spent on Website (minutes)") +
  ylab("Number of Pages Visited") +
  theme_minimal()
```

We apply gradient descent to learn optimal weights and predict whether future customers will buy by minimizing the loss function iteratively.

